Automated Formal Synthesis\\ of Safe Digital Controllers for Continuous Plants

### 2019-01-08 automatica/paper.tex
massive rewriting
should probably be split into separate commits, but i gave up

We provide experimental results showing that both our approaches are able to
efficiently synthesize safe controllers for a set of intricate physical
plant models taken from the digital control literature.
\addtodo{add more here once we have the experimental results ready}

@@

\note{Anticipate here connection with characteristic polynomial? See Section 4.}

\subsection{Symbolic execution}
We use the bounded model checker CBMC~\cite{ClarkeKL04} to model controller behaviour on digitial signals with finite
bit-precision.  We express these controller semantics as C programs, which is CBMC's preferred input language.  CBMC
symbolically executes every instruction in an input program by translating it to a boolean satisfiability problem.  At
the end of this process, the entire input program is translated to a boolean satisfiability formula which is satisfiable
iff a certain property about the program holds.  CBMC's API allows to specify these properties in the form of assertions
in the input program, and we use these properties to express that our generated controllers are modelled in a stable and
safe fashion. %\note{Do we need more details here? Should we explain e.g. SSA or give examples of how assignments, loops,
%etc. are turned into formulas? Part of Sec.~\ref{sec:errors} can probably be moved here.}

%-------------------------------
\subsection{Errors}
\label{sec:errors}

We account for the presence of errors originating from a number of sources, shown in Figure~\ref{fig:digitalsystem}:
\begin{itemize}
\item We represent the plant variables $x(t)$ and $u(t)$, which are continuous quantities, with finite precision in our model. 
which we do using standard interval arithmetic. 
\item The quantisation errors by the ADC, which converts the analogue signal $x(t)$ into the digital signal $x_k$, 
which is then fed into the controller cf. Figure~\ref{fig:digitalsystem}. 
\ifx\axelerator
We account for this error in the acceleration-based back-end by introducing a noise signal $\nu_1$ that over-approximates it. 
Further details are found in Section~\ref{sssec:abstraction}.  
In the multi-stage back-end we
\else
We
\fi
use bit-precise modelling in CBMC~\cite{ClarkeKL04}, a bounded model checker for C-programs, to accurately encompass this conversion. 
In other words, CBMC constructs a Boolean formula that explicitly unwinds the transition system for symbolically represented initial states. 
The values that the symbolically represented initial state may take are restricted by upper and lower bounds as described in section~\ref{}.
The Boolean formula constructed by CBMC models the ADC conversions precisely on the symbolically represented states at every time step. 
When we check for a safety property, the Boolean formula is satisfied only if there is an input state value for which the subsequent trace through the transition system eventually violates the safety property.

\item The representation errors due to performing finite-precision arithmetic in the controller block. 
\ifx\axelerator
In the acceleration-based back-end, 
this is again treated by the introduction of a noise signal that bounds the error. 
The multi-stage back-end uses 
\else
We use
\fi
the bit-precise modelling of finite-precision arithmetic in CBMC, 
in the same way as the ADC conversion was handled. 
We assume that the ADC represents numbers with a higher precision than the controller, 
and thus the precision is limited by the controller. This is a reasonable assumption based on commonly available hardware.
\item We assume that the DAC converts numbers to analogue signals with the same precision as the controller. 
As a result we can ignore any additional errors introduced by the DAC. Again, this assumption is reasonable based on commonly available hardware. 
\end{itemize}

\subsection{Notation for fixed- and floating-point precision}

In this paper we will use $\mathcal{F}_{\langle I,F \rangle}(x)$ to denote a real number $x$ expressed at a fixed-point precision, 
using $I$ bits to represent the integer part of the number and $F$ bits to represent the decimal part.

@@

There are a number of procedures that can be used for stability analysis of dynamical models~\cite{daes20161, Bessa16}, 
which can fit our automatic verification engine.  
%one based on Schur's decomposition
%and another one based on Jury's criterion~\cite{astrom1997computer}.  
Here we choose Jury stability criterion~\cite{astrom1997computer}, 
in view of its efficiency and ease of integration within our implementation.  
This method checks the stability working in the complex domain of the characteristic polynomial $S(z)$,  
%
% of the matrix $\left( \begin{array}{c} Y \\ U \end{array}\right)$, where $Y(z)$ is the closed-loop system output and U(z) is the controller output. 
%% \red{[unclear: what matrix? We need to clarify the relationship
%% between the char. poly. $S(z)$ and the later quantities $\Delta
%% N_{G}(z), \Delta D_{G}(z)$ and corresponding controller's
%% structure.]}.
%
considered in its general form as 
% for $S(z)$:
%
\begin{equation*}
S(z) = a_0z^N+a_1z^{N-1}+\cdots+a_{N-1}z+a_N,\,\, a_0\neq0. 
\end{equation*}
\note{The characteristic polynomial $S(z)$ represents the dynamical model in \ldots [clarify whether we use the CPol to describe the open- or closed-loop model. Relate parameters to matrices A,B,K. Notice that these parameters are used later in the Section where AA-based backend is explained.]}
%\cdsay{is this for the matrix $\left( \begin{array}{c} Y \\ U \end{array}\right)$?} 
% Assuming that $\hat{G}(z) = \frac{\hat{N_G}(z)}{\hat{D_G}(z)}$
% and $\hat{C}(z) = \frac{\hat{N_C}(z)}{\hat{D_C}(z)}$, then:
% $$S(z) = FWL[N_C(z)] \times \hat{N_G}(z) + FWL[D_C(z)] \times \hat{D_G}(z)$$
The following matrix $M$ with dimension $(2N-2)\times N$ and elements $m_{(\cdot),(\cdot)}$ is built from the coefficients of $S(z)$ as:  
%
$$
M=\left( 
\begin{array}{c}
V^{(0)}\\
V^{(1)}\\
\vdots\\
V^{(N-2)}
\end{array}
\right), 
$$
%
where $V^{(k)} = [v^{(k)}_{ij} ]_{2\times N}$ is such that:
%
$$
v_{ij}^{(0)}=\left\{
\begin{array}{ll}
a_{j-1}, & \mbox{if}~i=1\\
v_{(1)(N-j+1)}^{0},&\mbox{if}~i=2
\end{array}
\right.
$$
%
$$
v_{ij}^{(k)}=\left\{
\begin{array}{ll}
0,&\mbox{if}~j>n-k\\
v_{1j}^{(k-1)}-v_{2j}^{(k-1)} . \frac{v_{11}^{(k-1)}}{v_{21}^{(k-1)}}, & \mbox{if}~j\leq n-k ~\mbox{and}~i=1\\
v_{(1)(N-j+1)}^{k},& \mbox{if}~j\leq n-k ~\mbox{and}~i=2\\
\end{array}
\right.
$$
%
and where $k \in \mathbb{Z}$ is such that $0 < k < N - 2$.

@@

\begin{algorithm}[]
%\scriptsize
\begin{algorithmic}[1]
\Function{$\mathit{safetyCheck}()$}{}
\State assert($ \underline{u}  \leq u \leq \overline{u}$)
 \State set $x_0$ to be a \note{vertex state[?]}, e.g., $[\underline{x_0},\underline{x_0}]$	
\For {($c=0;~c < 2^\mathit{Num\_States};~c\mbox{++}$) }
  %\State assert($ \underline{x_0}  \leq x_0 \leq \overline{x_0}$)
	\For{($i=0;~i< k;~i\mbox{++}$)}
		%\State $u = (\langle I_p,F_p\rangle)((\langle I_c,F_c\rangle)K * (\langle I_c,F_c\rangle) x)$
		\State $u = (plant\_typet)((controller\_typet)K * (controller\_typet) x)$
		\State $x = A * x + B * u$
		\State assert($\underline{x} \leq x \leq \overline{x}$ )
  	\EndFor
  	\State set $x_0$ to be a new \note{vertex state}
  	\EndFor
  %\State Return $K$
\EndFunction
\end{algorithmic}
\caption{Safety check\label{alg:safetycheck}}
\end{algorithm}

\item The first {\sc verify} stage, {\sc safety}, shown in Alg.~\ref{alg:safetycheck} \note{$Num\_States = ?, plant\_typet = ? controller_typet =?$}, checks that the candidate
solution $K$, which we synthesized to be safe for at least one initial
state, is safe for \emph{all} possible initial states, i.e., does not reach
an unsafe state within $k$ steps.  After unfolding the transition system corresponding
to the previously synthesized controller $k$ steps, we check that the safety
specification holds for any initial state.

@@

\item The third {\sc verify} stage, {\sc complete}, checks that the current
$k$ is large enough to ensure safety for any $k'{>}k$.  Here, we compute the
completeness threshold $\overline{k}$ for the current candidate controller $K$ and
check that $k{\geq}\overline{k}$.  This is done by computing the number of time steps
required for the states to have completed a 360\textdegree circle, as illustrated in Fig.~\ref{fig:ct}. 
\note{This is only for spiralling dynamics - more generally, we need to clarify how the CT is computed. Shall we add the CAV17 proof?}

@@

As discussed in Section~\ref{sec:errors}, 
the models we consider have to account for errors from several sources: 
the numerical error introduced by the finite precision used to represent the plant (and its operations),  
the error introduced by the ADC/DAC conversions, 
and the error introduced by the limited precision used to represent the controller. 
First we consider the numerical errors introduced by the finite precision numbers (either fixed- or floating-point) used to model the plant and used by the digital controller. 
Then we account for the quantization errors introduced by the ADC/DAC.

\subsection{Errors due to numerical representations}

Let us first consider the case of fixed-point representation of the model, 
and of synthesis of fixed-point controllers.

Recall we use $\mathcal{F}_{\langle I,F \rangle}(x)$ denote a real number $x$
represented in a fixed point domain, with $I$ bits representing the integer
part and $F$ bits representing the decimal part.  The smallest number that
can be represented in this domain is $c_m=2^{-F}$.  Any mathematical
operations performed at the precision $\mathcal{F}_{\langle I,F \rangle}(x)$
will introduce errors, for which an upper bound can be
given~\cite{DBLP:conf/arith/BrainTRW15}:

@@

\ifx\axelerator
\subsection{ADC/DAC quantization errors modelled as noise} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
In this section we describe how quantization errors can be modelled as noise. 
Note that this is only applicable to the acceleration back end, as the multi-stage back 
end makes use of CBMC's bit-precise model checking capabilities to account 
for these deterministic errors.

@@

We separate our evaluation into two sets of results: fixed- and floating-point. 
For the majority of the {\it fixed-point} benchmarks, we observe that the abstraction-based
back-end is faster than the basic multi-staged verification approach; \textcolor{blue}{additionally, 
from 26 benchmarks, we observed that the abstraction-based back-end finds a solution for more benchmarks 
(14) than the multi-staged back-end (13)}.
%\note{we should clarify that this is out of 26 in total, not 191?}  
In direct comparison, the abstraction-based approach is on average able to find a
solution in approximately 53\% of the time required using the multi-staged
back-end, and has a median run-time 0.5\,s, which is twenty times smaller
than the multi-staged approach.  The two back-ends complement each other in terms of 
benchmark coverage, and together solve 58\% of the benchmarks \note{only? this does not sound much really}.  
%
For the majority of the benchmarks with {\it floating-point} implementation, 
we also observe that the abstraction-based back-end is much faster than the basic multi-staged verification approach, 
but both back-ends find the same amount of solutions (13).  
In direct comparison, the abstraction-based approach is on average able to find a
solution in approximately 50\% of the time required using the multi-staged
back-end, and has a median run-time 0.6\,s, which is forty times smaller
than the multi-staged approach.  The two back-ends complement each other in
benchmark coverage and together solve 54\% of the benchmarks \note{somewhat underwhelming}.

>>>>

We provide experimental results showing that we are able to
efficiently synthesize safe controllers for a set of intricate physical
plant models taken from the digital control literature.
%\addtodo{add more here once we have the experimental results ready}

@@

The gain matrix $K$ can be set so that the closed-loop discrete dynamics are
shaped as desired, for instance according to a specific stability goal or
around a specific dynamical behavior \cite{astrom1997computer}.  As argued
later in this work, we will target more complex objectives, such as
quantitative safety requirements, which are not typical in the digital
control literature.  Further, we will embrace the digital nature of the
controller, which manipulates quantized signals as discrete quantities represented with finite precision.

\subsection{Stability of closed-loop systems}

In this work we employ asymptotic stability in the CEGIS loop,  
as an objective for guessing controllers that are later proven sound over safety requirements.  
Asymptotic stability is a property that amounts to convergence of the model executions to an equilibrium point, 
starting from any states in a neighborhood of the point (see Figure~\ref{fig:ct} for the portrait of a stable execution, converging to the origin).   
In the case of linear systems as in~\eqref{eq:closedloopss}, 
considered with a zero reference signal, 
the equilibrium point of interest is the origin.

A discrete-time LTI system as \eqref{eq:closedloopss} is asymptotically
stable if all the roots of its characteristic polynomial (i.e., the
eigenvalues of the closed-loop matrix $A_d - B_d K$) are inside the unity
circle of the complex plane, i.e., their absolute values are strictly less than
one~\cite{astrom1997computer} (this simple sufficient condition can be generalised, 
however this is not necessary in our work).  
In this paper, we express this stability specification $\phi_\mathit{stability}$ in terms of a check known as
\emph{Jury's criterion}~\cite{fadali}: this is an easy algebraic formula to
select the entries of matrix $K$ so that the closed-loop dynamics are shaped
as desired.

\subsection{Safety of closed-loop systems}

We are not limited to the synthesis of digital stabilizing controllers -- a
well known task in the literature on digital control systems -- but target
safety requirements with an overall approach that is sound and automated. 
More specifically, we require that the closed-loop system
\eqref{eq:closedloopss} meets given safety specifications.  A safety
specification gives raise to a requirement on the states of the model, such
that the feedback controller (namely the choice of the gains matrix~$K$)
must ensure that the state never violates the requirement.  Note that a
stable, closed-loop system is not necessarily a safe system: indeed, the
state values may leave the safe part of the state space while they converge
to the equilibrium, which is typical in the case of oscillatory dynamics. 
%However, a safe system will always be stable.  
In~this work, the safety property is expressed as:
%
\begin{equation}
\label{eq:safetyliteral}
\phi_\mathit{safety}\iff \forall k\ge 0.\, \bigwedge_{i=1}^{n}{\underline{x_{i}} \leq x_{i,k} \leq \overline{x_{i}}},
\end{equation}
%
%\addtodo{We consider $x_{i}^{-}$ and $x_{i}^{+}$ to be -1 and 1, respectively.}
%
where $\underline{x_{i}}$ and $\overline{x_{i}}$ are lower and upper bounds
for the $i$-th coordinate $x_{i}$ of state $x\in \mathbb R^n$ at the $k$-th
instant, respectively.  This means that the states will always be within an $n$-dimensional hyper-box.

Furthermore, it is practically relevant to consider the 
constraints $\phi_\mathit{input}$ on the input
signal $u_{k}$ and $\phi_\mathit{init}$ on the initial states $x_0$,
which we assume have given bounds:
$\phi_\mathit{input} = {\forall k.\underline{u} \leq u_{k} \leq \overline{u}} $, 
$\phi_\mathit{init} = \bigwedge_{i=1}^{n} \underline{x_{i,0}} \leq x_{i,0} \leq \overline{x_{i,0}}.$
For the former, this means that the control input might saturates in view of
physical constraints.

\subsection{Bounded Model Checking}
We use the bounded model checker CBMC~\cite{ClarkeKL04} to model controller behaviour on digitial signals with finite
bit-precision.  
We express these controller semantics as C programs, which is CBMC's preferred input language.  
CBMC
symbolically executes every instruction in an input program by translating it to a boolean satisfiability problem.  
At
the end of this process, the entire input program is translated to a boolean satisfiability formula which is satisfiable
iff a certain property about the program holds.  
CBMC's API allows to specify these properties in the form of assertions
in the input program, and we use these properties to express that our generated controllers are stable and safe.
%\note{Do we need more details here? Should we explain e.g. SSA or give examples of how assignments, loops,
%etc. are turned into formulas? Part of Sec.~\ref{sec:errors} can probably be moved here.}

%-------------------------------
\subsection{Finite precision errors}
\label{sec:errors}

The key contribution of our work is that, when synthesising controllers, we precisely model the finite-precision arithmetic in the digital controller, guaranteeing that the controllers are safe. The finite precision arithmetic introduces the following errors, shown in Figure~\ref{fig:digitalsystem}. These errors are present in the real world, and we must model them accurately 
in order to capture the full behaviour of the system. Details of how we model this arithmetic are in Section~\ref{sec:}:

\begin{itemize}

\item The ADC introduces quantisation errors when it converts the analogue signal $x(t)$ into the digital signal $x_k$. This signal is then fed into the controller cf. Figure~\ref{fig:digitalsystem}.

\item The controller block performs arithmetic at a finite-precision, which introduces additional errors on top of the error introduced by the ADC. We assume the ADC represents numbers with at least the same precision
as the controller and thus the precision is limited by the controller. This is a reasonable assumption based on commonly available hardware.

\item The DAC could potentially introduce further errors. However, we assume that the DAC converts numbers to analogue signals with the same precision as the controller. 
As a result we can ignore any additional errors introduced by the DAC. Again, this assumption is reasonable based on commonly available hardware. 
\end{itemize}

The arithmetic and representations used in our modelling are also carried out at a finite precision, and so we must also account for the errors in our own modelling. Specifically our model of the plant variables  $x(t)$ and $u(t)$, which are, in reality, continuous quantities, with finite precision in our model. We guarantee that the precision we use to represent the plant variables is at least as precise as the precision used in the digital controller and we use interval arithmetic to bound the errors, as detailed in Section~\ref{sec:}.

\subsection{Notation for fixed- and floating-point precision}  
In this paper we will use $\mathcal{F}_{\langle I,F \rangle}(x)$ to denote a real number $x$ expressed at a fixed-point precision, 
using $I$ bits to represent the integer part of the number and $F$ bits to represent the decimal part.

@@

There are a number of procedures that can be used for stability analysis of dynamical models~\cite{daes20161, Bessa16}, 
which can fit our automatic verification engine.  
%one based on Schur's decomposition
%and another one based on Jury's criterion~\cite{astrom1997computer}.  
Here we choose Jury stability criterion~\cite{astrom1997computer}, 
in view of its efficiency and ease of integration within our implementation.  
This method checks the stability working in the complex domain of the characteristic polynomial $S(z)$,  
%
% of the matrix $\left( \begin{array}{c} Y \\ U \end{array}\right)$, where $Y(z)$ is the closed-loop system output and U(z) is the controller output. 
%% \red{[unclear: what matrix? We need to clarify the relationship
%% between the char. poly. $S(z)$ and the later quantities $\Delta
%% N_{G}(z), \Delta D_{G}(z)$ and corresponding controller's
%% structure.]}.
%
considered in its general form as 
% for $S(z)$:
%
\begin{equation*}
S(z) = a_0z^N+a_1z^{N-1}+\cdots+a_{N-1}z+a_N,\,\, a_0\neq0. 
\end{equation*}
An LTI system is stable if all the roots of its characteristic polynomial $S(z)$ (i.e., the 
eigen values of the closed loop matrix $A_d-B_dK$) are inside the unit circle of the complex plane, 
i.e., their absolute values are less than one.

%\note{The characteristic polynomial $S(z)$ represents the dynamical model in \ldots [clarify whether we use the CPol to describe the open- or closed-loop model. Relate parameters to matrices A,B,K. Notice that these parameters are used later in the Section where AA-based backend is explained.]}

The following matrix $M$ with dimension $(2N-2)\times N$ and elements $m_{(\cdot),(\cdot)}$ is built from the coefficients of $S(z)$ as:  
%
$$
M=\left( 
\begin{array}{c}
V^{(0)}\\
V^{(1)}\\
\vdots\\
V^{(N-2)}
\end{array}
\right), 
$$
%
where $V^{(k)} = [v^{(k)}_{ij} ]_{2\times N}$ is such that:
%
$$
v_{ij}^{(0)}=\left\{
\begin{array}{ll}
a_{j-1}, & \mbox{if}~i=1\\
v_{(1)(N-j+1)}^{0},&\mbox{if}~i=2
\end{array}
\right.
$$
%
$$
v_{ij}^{(k)}=\left\{
\begin{array}{ll}
0,&\mbox{if}~j>n-k\\
v_{1j}^{(k-1)}-v_{2j}^{(k-1)} . \frac{v_{11}^{(k-1)}}{v_{21}^{(k-1)}}, & \mbox{if}~j\leq n-k ~\mbox{and}~i=1\\
v_{(1)(N-j+1)}^{k},& \mbox{if}~j\leq n-k ~\mbox{and}~i=2\\
\end{array}
\right.
$$
%
and where $k \in \mathbb{Z}$ is such that $0 < k < N - 2$.

@@

\begin{algorithm}[]
%\scriptsize
\begin{algorithmic}[1]
\Function{$\mathit{safetyCheck}()$}{}
\State assert($ \underline{u}  \leq u \leq \overline{u}$)
 \State set $x_0$ to be a vertex state, e.g., $[\underline{x_0},\underline{x_0}]$	
\For {($c=0;~c < 2^\mathit{Num\_States};~c\mbox{++}$) }
  %\State assert($ \underline{x_0}  \leq x_0 \leq \overline{x_0}$)
	\For{($i=0;~i< k;~i\mbox{++}$)}
		%\State $u = (\langle I_p,F_p\rangle)((\langle I_c,F_c\rangle)K * (\langle I_c,F_c\rangle) x)$
		\State $u = (plant\_typet)((controller\_typet)K * (controller\_typet) x)$
		\State $x = A * x + B * u$
		\State assert($\underline{x} \leq x \leq \overline{x}$ )
  	\EndFor
  	\State set $x_0$ to be a new vertex state
  	\EndFor
  %\State Return $K$
\EndFunction
\end{algorithmic}
\caption{Safety check\label{alg:safetycheck}}
\end{algorithm}

\item The first {\sc verify} stage, {\sc safety} is shown in Alg.~\ref{alg:safetycheck}. 
The algorithm checks that the candidate
solution $K$, which we synthesized to be safe for at least one initial
state, is safe for \emph{all} possible initial states, i.e., does not reach
an unsafe state within $k$ steps.  After unfolding the transition system corresponding
to the previously synthesized controller $k$ steps, we check that the safety
specification holds for any initial state.

We use $(controller\_typet)$ to denote a
cast or conversion to the controller precision, and $(plant\_typet)$ to denote a cast or conversion to the 
plant precision. A \emph{vertex state} is defined as a state where all values are either equal to the upper or lower
bound of the initial states. It is sufficient to verify that the controller is safe for all vertex
states, and the proof is shown later in this section. We thus only need to iterate through the $2^{Num\_States}$ vertex
states, where $Num\_States$ is the number of dimensions in $x$.

\begin{theorem} 
If a controller is safe for each of the corner cases of our hypercube of
allowed initial states, i.e., the vertex states, it is safe for any initial state in the hypercube. 
\end{theorem}

Thus we only need to check $2^n$ initial states, where $n$
is the dimension of the state space (number of continuous variables). 
\begin{proof}
Consider the set of initial states, $X_0$, which we assume to be convex since it is a hypercube. 
Name $v_i$ its vertexes, where $i=1,\ldots, 2^n$.  
Thus any point $x \in X_0$ can be expressed by convexity as $x = \sum_{i=1}^{2^n} \alpha_i v_i$, 
where $\sum_{i=1}^{2^n} \alpha_i =1$. Then if $x_0=x$, we obtain 
\begin{align*}
x_k   &= (A_d - B_d K)^k x \\
&= (A_d - B_d K)^k \sum_{i=1}^{2^n} \alpha_i v_i \\
      &= \sum_{i=1}^{2^n} \alpha_i (A_d - B_d K)^k v_i \\
      &= \sum_{i=1}^{2^n} \alpha_i x_k^i, 
 \end{align*}     
%
where $x_k^i$ denotes the trajectories obtained from the single vertex
$v_i$.  We~conclude that any $k$-step trajectory is encompassed, within a
convex set, by those generated from the vertices. 
\qed
\end{proof}

@@

\item The third {\sc verify} stage, {\sc complete}, checks that the current
$k$ is large enough to ensure safety for any $k'{>}k$.  Here, we compute the
completeness threshold $\overline{k}$ for the current candidate controller $K$ and
check that $k{\geq}\overline{k}$.  This is done by computing the number of time steps
required for the states to have completed a 360\textdegree circle, as illustrated in Fig.~\ref{fig:ct}.

\begin{theorem} There exists a finite $\overline{k}$ such that it is
sufficient to unwind the closed-loop state-space system up to $\overline{k}$
in order to ensure that $\phi_\mathit{safety}$ holds. 
\end{theorem}

\begin{proof}
%
A stable control system is known to have converging dynamics.  Assume the
closed-loop matrix eigenvalues are not repeated (which is sensible to do,
since we select them).  The distance of the trajectory from the reference
point (origin) decreases over time within subspaces related to real-valued
eigenvalues; however, this is not the case in general when dealing with
complex eigenvalues.  Consider the closed-loop matrix that updates the
states in every discrete time step, and select the eigenvalue $\vartheta$
with the smallest (non-trivial) imaginary value.  Between every pair of
consecutive time steps $k\,T_s$ and $(k+1)\,T_s$, the dynamics projected on
the corresponding eigenspace rotate $\vartheta T_s$ radians.  Thus, taking
$\overline{k}$ as the ceiling of $\frac{2\pi}{\vartheta T_s}$, after
$k{\geq}\overline{k}$ steps we have completed a full rotation, which results
in a point closer to the origin.  The synthesized $\overline{k}$ is the completeness threshold.
\qed 
%
\end{proof}

@@

As discussed in Section~\ref{sec:errors}, 
the models we consider must account for errors from several sources: 
we must bound the numerical error introduced by the finite precision we use to represent the plant (and its operations), and precisely model
the error introduced by the ADC/DAC conversions
and the error introduced by the limited precision used to represent the controller. 
We use interval arithmetic to bound the error introduced by the finite precision plant, and bounded model checking to precisely model the error introduced by the ADC/DAC and the finite precision controller.

\subsection{Bit-precise symbolic model checking}
As described in section~\ref{sec:preliminaries}, we use CBMC, a bit-precise bounded model checker to synthesise and verify candidate controllers. CBMC performs precisely the fixed or floating point arithmetic used in the controller, as well as the ADC/DAC conversions, according to the IEEE standards.

\subsection{Interval arithmetic over errors in numeric representation} 
We use finite-precision arithmetic to model the plant. This is an approximation that speeds up each CEGIS
iteration but necessitates a further stage where we verify that the errors introduced by the
approximation have not caused us to generate a controller that is unsafe when executed with real numbers 
for the plant. In this stage, we represent the plant using double precision floating point
and we use the Boost interval arithmetic library~\cite{} to bound the error in this representation. We use a compositional numeric library to perform the fixed point arithmetic for the controller~\cite{}. We check the controller is safe starting from each vertex state, which we have shown to be sufficient to prove safety from any initial state in the hypercube of possible initial states.

We outline here the mathematics behind bounding the errors on the double precision floating point.
Recall we use $\mathcal{F}_{\langle E,M \rangle}(x)$ denote a real number $x$
represented in a floating point domain, with $E$ bits representing the exponent and $M$ bits representing the mantissa part. Any representation of a real number using the floating point domain will
introduce an error, for which an upper bound can be given~\cite{DBLP:conf/arith/BrainTRW15}.
For each number represented in floating point, we store an interval that encompasses this error.
Further
mathematical
operations performed at the precision $\mathcal{F}_{\langle E,M \rangle}(x)$
will propagate these errors forward and 
introduce further errors for which bounds can be derived~\cite{DBLP:conf/arith/BrainTRW15}.

The fixed point arithmetic of the digital controller is performed on the upper and lower limit of each interval independently. Since the precision of the floating point is greater than the precision
of the controller, this is guaranteed to bound the real behaviour of the controller.

\ifx\axelerator

@@

\subsection{ADC/DAC quantization errors modelled as noise} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
In this section we describe how quantization errors can be modelled as noise. 
Note that this is only applicable to the acceleration back end, as the multi-stage back 
end makes use of CBMC's bit-precise model checking capabilities to account 
for these deterministic errors.

@@

We separate our evaluation into two sets of results: fixed- and floating-point. 
%For the majority of the {\it fixed-point} benchmarks, we observe that the abstraction-based
%back-end is faster than the basic multi-staged verification approach; \textcolor{blue}{additionally, 
%from 26 benchmarks, we observed that the abstraction-based back-end finds a solution for more benchmarks 
%(14) than the multi-staged back-end (13)}.
%%\note{we should clarify that this is out of 26 in total, not 191?}  
%In direct comparison, the abstraction-based approach is on average able to find a
%solution in approximately 53\% of the time required using the multi-staged
%back-end, and has a median run-time 0.5\,s, which is twenty times smaller
%than the multi-staged approach.  The two back-ends complement each other in terms of 
%benchmark coverage, and together solve 58\% of the benchmarks \note{only? this does not sound much really}.  
%%
%For the majority of the benchmarks with {\it floating-point} implementation, 
%we also observe that the abstraction-based back-end is much faster than the basic multi-staged verification approach, 
%but both back-ends find the same amount of solutions (13).  
%In direct comparison, the abstraction-based approach is on average able to find a
%solution in approximately 50\% of the time required using the multi-staged
%back-end, and has a median run-time 0.6\,s, which is forty times smaller
%than the multi-staged approach.  The two back-ends complement each other in
%benchmark coverage and together solve 54\% of the benchmarks \note{somewhat underwhelming}.

%\begin{table*}
%\centering
%\begin{tabular}{| r | l | c | c | r | c | r | c | r | c | r |}
%%
%\hline
%\# & \multicolumn{1}{|c|}{Benchmark} & \multicolumn{1}{|c|}{Order} & \multicolumn{4}{|c|}{Multi-staged}                 & \multicolumn{4}{|c|}{Abstraction} \\
%   &                                  & & \multicolumn{1}{|c|}{$\mathcal{F}_{\langle I_p,F_p \rangle}$} & \multicolumn{1}{|c|}{Time} & \multicolumn{1}{|c|}{$\mathcal{F}_{\langle E_p, M_p \rangle}$} & \multicolumn{1}{|c|}{Time} & \multicolumn{1}{|c|}{$\mathcal{F}_{\langle I_p,F_p \rangle}$} & \multicolumn{1}{|c|}{Time} & \multicolumn{1}{|c|}{$\mathcal{F}_{\langle E_p, M_p \rangle}$} & \multicolumn{1}{|c|}{Time}\\\hline
%1  & Acrobot     & 4 & 16,8 & 61959.4\,s &  & ~\xmark & & ~\xmark &  & ~\xmark \\
%2  & Antenna     & 6 &  & ~\xmark &  & ~\xmark & & ~\xmark & & 2903s~\xmark\\
%3  & Ballmaglev  & 3 &  & ~\xmark &  & ~\xmark & & ~\xmark & & ~\xmark\\
%4  & Bioreact    & 3 & 16,8 & 11.5\,s & 16,10 &14.6\,s & 24,14 & 0.4\,s & 16,10 & 0.5\,s\\
%5  & Chen\_ex1   & 3 & 16,8 & 10.3\,s  & 16,10 & 9.2\,s & & ~\xmark & 16,10 & 1.4\,s\\
%6  & Chen\_ex2   & 8& & ~\xmark & 32,23 & 4683.8\,s &  & ~\xmark & & ~\xmark\\
%7  & Chen\_ex3   & 7 &   & ~\xmark & & ~\xmark & & ~\xmark & & ~\xmark\\
%8  & Chen\_ex4   & 9 & & ~\xmark & & ~\xmark & & ~\xmark & & ~\xmark\\
%9  & Cruise      & 1 & 16,8 & 9.6\,s & 16,10 & 6.5\,s & 24,14 & 0.3\,s& 16,10 & 0.3\,s\\
%10 & CruiseHSCC & 1 & 16,8 & 9.6\,s & 16,10 & 6.6\,s &  24,14 & 0.3\,s & 16,10 & 0.3\,s \\
%11 & Cstr & 3 & & ~\xmark & & ~\xmark & & ~\xmark & & 181s~\xmark\\
%12 & Cstrtmp  & 3 &  & ~\xmark & & ~\xmark & 24,14 & 0.5\,s & & ~\xmark\\
%13 & DCmotor   & 2 & 16,8  & 11.3\,s & 16,10 & 7.8\,s & 24,14 & 0.4\,s & 16,10 & 0.5\,s (4.0\,s) \\
%14 & Flexbeam   & 6 & & ~\xmark & & ~\xmark & 24,14 & 72.1\,s & & ~\xmark\\
%15 & Guidance chen  & 5 & & ~\xmark & & ~\xmark & & ~\xmark & & ~\xmark\\
%16 & Helicopter   & 3 & 16,8 & 13.8\,s & 32,23 & 148\,s & 24,14 & 0.8\,s& 16,10 & 1.1\,s (56.6\,s)\\
%17 & Invpendulum\_cartpos & 4 & & ~\xmark & & ~\xmark & & ~\xmark & & ~\xmark\\
%18 & Invpendulum\_pendang & 2 & 24,12 & 9.8\,s & 16,10 & 28.2\,s & 24,14 & 0.4\,s & 16,10 & 0.5\,s\\
%19 & Magpointer   & 2 & 16,8 & 10.1\,s & 16,10 & 20.4\,s & 24,14  & 1\,s& 16,10 & 1.3\,s (\xmark)\\
%20 & Magsuspension  & 2 & 12,12  & 22.2\,s  & 16,10 & 34.8\,s & 24,14 & 0.5\,s & 16,10 & 0.6\,s (\xmark)\\
%21 & Pendulum   & 2 & 16,8 & 8.5\,s & 32,23 & 24\,s & 24,14 & 0.4\,s & 16,10 & 0.5\,s (2.4\,s)\\
%22 & Regulator   & 5 & & ~\xmark & & ~\xmark & &~\xmark  & & 529s~\xmark\\
%23 & Springmassdamper & 2 & & ~\xmark & & ~\xmark & & ~\xmark & & 14.1s~\xmark\\
%24 & streamdrum   & 3 & 16,8  & 332.5\,s & 16,10 & 590.5\,s & 24,14 & 0.7\,s & 16,10 &1.3\,s\\
%25 & Suspension  & 4 & 16,8  & 17.2\,s & 16,10 &1373\,s & 56,30 & 15.3\,s & 16,10 & 4.4\,s (313.6\,s)\\
%26 & Tapedriver   & 3 & 16,8  & 8.2\,s & & ~\xmark & 64,22 & 2.1\,s & 16,10 & 1.1\,s (6.0\,s)\\
%\hline
%%
%\end{tabular}
%\vspace{0.05in}
%\caption{Experimental results\label{tab:results}}
%\end{table*}
